Topological Eulerian Synthesis of Slow Motion Periodic Videos

### 2018-01-30 Paper/paper.tex
Finished code for batch cluster robustness tests
Major reworking of the intro of the paper, fixing block diagram, and adding an XT slice for the fan video

\title{Eulerian Synthesis of Slow Motion Periodic Videos}

@@

\begin{abstract}
Abstract
\end{abstract}
%
\begin{keywords}
One, two, three, four, five
\end{keywords}
%

\section{Introduction}

Repetitive, periodic motions are ubiquitous in our world, such as animal locomotion (walking/flapping/slithering/swimming), mechanical motions (wheels spinning, pistons oscillating), biological rhythms (heartbeat, breathing), musical rhythms (drumming, feet tapping), etc.  In this work, we consider videos of periodic motion.  One avenue of interest is to analyze these videos in fine detail, such as for characterizing subtle progressions of blood flow in the face during a heartbeat for medical diagnostic purposes.  It may also be important to understand and visualize variations from cycle to cycle, such as in a repetitive automated action on an assembly line, where too much variation could indicate the onset of failure, or in athletic activities for optimizing performance.

Our approach is {\em Eulerian}; that is, we process the video pixel by pixel with no tracking.  Unlike Fourier-based Eulerian motion video processing tools (\cite{wu2012eulerian, wadhwa2013phase}), however, we take a {\em geometric} approach by quantifying {\em sliding window embeddings} of our videos (Section~\ref{sec:slidingwindow}).  Sliding window embeddings of periodic time series form samples of a topological loop \cite{perea2015sliding}, as do generalized multivariate sliding windows on periodic video data \cite{traliehigh, tralie2017quasi}.  This implies that, at the right scale, a nearest neighbor graph built on the sliding window point cloud should be approximately a circle graph, even if the overall extrinsic geometry is quite complicated.  Bearing this in mind, we use Laplacian eigenmaps \cite{belkin2003laplacian} of this embedding to construct maps from these point clouds to the circle, or {\em circular coordinates} (Section~\ref{sec:laplacian}).  In particular, each sliding window is assigned a phase $\phi \in [0, 2\pi]$.  To autotune the scale at which Laplacian eigenmaps are built, we use topological data analysis \cite{edelsbrunner2000topological,edelsbrunner2008persistent,edelsbrunner2010computational,carlsson2009topology,ghrist2014elementary} to automatically find the scale at which the most prominent loop in the data occurs.

Once we have extracted a phase for each window, we use this information to reorder the content of the video.  We also exploit the sliding window structure to have different windows vote on the final pixels in each frame of the template (Section~\ref{sec:cyclereordering}).  As an illustration, Figure~\ref{fig:ConceptFigure} shows an example of our technique on a sampled periodic time series (i.e. single pixel grayscale video).  There are only 12 samples per period in the original signal, so the details of each period coarse and noisy.  However, once they are re-sorted, we get a nice, fine-detailed representation of one period.  The extension of this technique to videos will be described in detail in Section~\ref{sec:methods}.

\section{Releated Work}
Our technique unifies many disparate ideas that have appeared before in different contexts, in addition to the references we mentioned in the introduction.  Sliding window embeddings, or ``delay reconstructions,'' have found a diverse array of applications in activity recognition \cite{frank2010activity,venkataraman2016shape}, gene expression data \cite{perea2015sw1pers}, EEG analysis \cite{stam2005nonlinear, plesnik2014detection}, audio and music analysis \cite{herzel1994analysis,serra2009cross,bello2011measuring,traliemoebius}, video analysis \cite{schodl2000video, tralie2017quasi}, and motion capture analysis \cite{venkataraman2016shape}.  All of these works focus on analysis rather than synthesis, as we do, but there has been work using sliding window embeddings to synthesize ``video textures,'' or videos that repeat dynamics indefinitely \cite{schodl2000video}, which are similar in spirit to our infinitely looping periodic representatives.  The graph Laplacian, on the other hand, has been used to re-arrange images around a loop as a pre-processing step for structure from motion \cite{averbuch2015ringit}, and vector diffusion maps have been used to re-order the frames of microscope images a developing embryo around a loop \cite{dsilva2015diffusionvecordering}.  A different topological approach with cohomology circular coordinates\cite{de2011persistent} was used to parameterize a sliding window embedding of the Lorenz attractor \cite{de2012topological} and periodic activites in motion capture data \cite{vejdemo2015cohomological}.  Neural networks have also been trained to find circular coordinates \cite{levy2015live,anafi2017cyclops}.  Finally, perhaps the most related to our goal is work synthesizing seamless video loops by clustering pixels into different periods and adjusting them to loop through a common period \cite{Liao2013VideoLoops,Liao2015VideoLoops}.  TODO: talk about applications of TDA?

\section{Methods}
\label{sec:methods}

@@

We now describe our algorithm in several stages.  Throughout our discussion, we assume we are dealing with 3 channel color videos.  For a frame resolution of $W \times H$, each frame can be thought of as a Euclidean vector in $\mathbb{R}^{W \times H \times 3}$.  For a video with $N$ frames, we flatten each frame to a 1D vector and stack them row wise along an $N \times 3WH$ matrix $X$.  Whenever we perform a linear operation on all of the pixels at once, we can use the SVD $X = USV^T$ to reduce memory and computational costs by working with $US$ in place of $X$, which is in $N$ dimensions instead of $3WH \gg N$ (similar processing was done in \cite{turk1991eigenfaces} and \cite{tralie2017quasi}).  The only nonlinear operation is the median operation (Section~\ref{sec:cyclereordering}), and we project back into pixel space using $V^T$ before taking the median. Algorithm~\ref{alg:videoreordering} shows the full details of our method in pseudocode.

@@

where $\tau$ is a step size and $d$ is the dimension of the embedding.  In this work, we fix $\tau = 1$ for simplicity and drop $\tau$ in the subscript, and we also sample times $t$ only at integer frame indices.  As shown by Takens \cite{takens1981detecting}, a sliding window of dimension $d = 2m+1$ of even a single generic observation function (time series) of a dynamical system of intrinsic dimension $m$ is sufficient to reconstruct a topological embedding of the underlying trajectory in the original state space.  For periodic signals, this state space is a torus, and $d$ should be twice the number of harmonics present to injectively reconstruct loops on that torus \cite{perea2015sliding}.  The same is true of videos in general \cite{tralie2017quasi}.  Furthermore, the sliding window length $(d-1) \tau$, or simply $d-1$ in our case, maximizes the roundness of the embeddings when $d-1 = k T$ for some integer $k$, where $T$ is the fundamental period.  Since we use topological data analysis (TDA) to find the scale of the loop, and since TDA measures roundness, it is important that we tune our window size to be an integer multiple of the period.  To do this, we perform 1D ISOMAP \cite{tenenbaum2000global} on the original frames of the video to generate a 1D ``surrogate signal,'' and we perform an autocorrelation-based fundamental frequency estimation \cite{Mcleod05asmarter} on this signal to estimate the period and, hence, the appropriate $d$.  This is quite similar to the diffusion maps + autocorrelation approach in \cite{tralie2017quasi} for analyzing periodic videos.  This estimation often returns multiples of the fundamental period, which is fine for our purposes.

@@

@@

@@

\begin{figure}[h!]
\centering
\includegraphics[width=0.8\columnwidth]{XTSlice.pdf}
\caption{An XT slice of a line of pixels (magenta line, upper left) over time for the original video and for reordered videos with and without median consensus.}
\label{fig:XTSlice}
\end{figure}

Figure~\ref{fig:XTSlice} shows the difference between a simple reordering and a median consensus reordering.  Due to natural variation from cycle to cycle, the simple reordering has many temporal discontinuities when interleaving these cycles.  By contrast, the median voting is clean, and it has the added benefit of removing nonperiodic background components.

@@

Interpolated windows can be projected back to pixel space one by one to save memory (Line~\ref{algline:medianconsensus} of Algorithm~\ref{alg:videoreordering})

@@

Robustness experiments, jumping jacks 2 men, two exercise videos from \cite{levy2015live}, beating heart videos from \cite{traliehigh}, \cite{wu2012eulerian}, and \cite{wadhwa2013phase}.

>>>>

\title{Topological Eulerian Synthesis of Slow Motion Periodic Videos}

@@

\begin{abstract}

The majority of our pipeline is fully automated, using tools from topological data analysis in concert with nonlinear dimenison reduction to autotune the spatial and temporal scales of the period.

\end{abstract}
%
\begin{keywords}
One, two, three, four, five
\end{keywords}
%

\section{Introduction}
%Background, drift, noise, large motion, static pixels, scale finding, overall automation [fundamental frequency, etc].  Sliding window "time regularization"

Repetitive, periodic motions are ubiquitous in our world, such as animal locomotion (walking/flapping/slithering/swimming), mechanical motions (wheels spinning, pistons oscillating), biological rhythms (heartbeat, breathing), musical rhythms (drumming, feet tapping), etc.  In this work, we consider the problem of synthesizing a fine detail, ``slow motion'' template cycle of motion from a video of such a motion with many periods.  This fine detailed analysis can be applied, for example, to characterize subtle progressions of blood flow in the face during a heartbeat \cite{kumar2015distanceppg}.  Having access to a consensus template can also be used to visualize variations from cycle to cycle, such as in a repetitive automated action on an assembly line, where too much variation could indicate the onset of failure (TODO: citation), or in athletic activities for optimizing performance.

Our approach to slow motion templates is {\em Eulerian}; that is, we process the video pixel by pixel with no tracking.  Eulerian approaches for video synthesis are attractive due to their ease of implementation.  For instance, a simple Fourier bandpass filtering and amplifying pixels can successfully elucidate subtle periodic motions in videos \cite{wu2012eulerian, wadhwa2013phase}.  Additionally, time domain Eulerian approaches have been used to synthesize infinitely playing ``video textures'' using random walks \cite{schodl2000video}, or to synthesize perfectly looping video templates after grouping pixels with similar periods together and devising a common period for all groups \cite{Liao2013VideoLoops,Liao2015VideoLoops}.  There are some pitfalls with Eulerian techniques, however.  Purely Eulerian amplification techniques are known to fail for large motions, and they also require the user to manually specify frequencies of interest \cite{wu2012eulerian, wadhwa2013phase}.  Furthermore, drift from period to period due to shaking and varying motions from cycle to cycle can present a problem, as can unrelated motion in the background \cite{stauffer1999adaptive}.

To address the problems with an Eulerian representation, we devise a geometric/topological Eulerian framework byconstructing {\em sliding window embeddings} of our videos (Section~\ref{sec:slidingwindow}).  Sliding window embeddings, or ``delay reconstructions,'' have found a diverse array of applications in activity recognition \cite{frank2010activity,venkataraman2016shape}, gene expression data \cite{perea2015sw1pers}, EEG analysis \cite{stam2005nonlinear, plesnik2014detection}, audio and music analysis \cite{herzel1994analysis,serra2009cross,bello2011measuring,traliemoebius}, video analysis \cite{tralie2017quasi}, and motion capture analysis \cite{venkataraman2016shape}.  Sliding window embeddings of periodic time series, in particular, form samples of a topological loop \cite{perea2015sliding}, as do generalized multivariate sliding windows on periodic video data \cite{traliehigh, tralie2017quasi}.  We autotune the sliding window size using fundamental frequency estimation \cite{Mcleod05asmarter}, and we use topological data analysis \cite{edelsbrunner2000topological,edelsbrunner2008persistent,edelsbrunner2010computational,carlsson2009topology,ghrist2014elementary} to find the appropriate spatial scale of the loop.  We then parameterize the video by assigning each window a phase $\phi \in [0, 2\pi]$ using Laplacian Eigenmaps \cite{belkin2003laplacian}.  We then re-order the sliding windows by phase and exploit the sliding window structure to have different windows vote on the final pixels in each frame of the template, which helps with drift (Section~\ref{sec:cyclereordering}).  As an illustration, Figure~\ref{fig:ConceptFigure} shows an example of our technique on a sampled periodic time series (i.e. single pixel grayscale video).  There are only 12 samples per period in the original signal, so the details of each period coarse and noisy.  However, once they are re-sorted, we get a nice, fine-detailed representation of one period.  The extension of this technique to videos will be described in detail in Section~\ref{sec:methods}.

Many prior works use a combination of sliding window embeddings and topological data analysis, such as detecting the Lorenz attractor \cite{de2012topological}, quantifying circadian rhythms in gene expression data \cite{perea2015sliding}, detecting chatter in mechanical systems \cite{khasawneh2016chatter}, detecting wheezing in audio \cite{emrani2014real}, quantifying periodic activites in motion capture data \cite{vejdemo2015cohomological, venkataraman2016persistent}, quantifying vocal fold anomalies in video \cite{tralie2017quasi}.  As shown by the authors of \cite{tralie2017quasi}, sliding window embeddings on Eulerian videos provide a sort of ``time regularization,'' which can mitigate the effects of drift, leading to a cleaner loop,

%The graph Laplacian, on the other hand, has been used to re-arrange images around a loop as a pre-processing step for structure from motion \cite{averbuch2015ringit}, and vector diffusion maps have been used to re-order the frames of microscope images a developing embryo around a loop \cite{dsilva2015diffusionvecordering}.  A different topological approach with cohomology circular coordinates\cite{de2011persistent} was used to parameterize a sliding window embedding of the .  Neural networks have also been trained to find circular coordinates \cite{levy2015live,anafi2017cyclops}.

\section{Slow Motion Templates}
\label{sec:methods}

@@

We now describe our algorithm in several stages.  Throughout our discussion, we assume we are given 3 channel color videos with $W \times H$ resolution frames, so that each frame is a Euclidean vector in $\mathbb{R}^{W \times H \times 3}$.  For a video with $N$ frames, we flatten each frame to a 1D vector and stack them row wise along an $N \times 3WH$ matrix $X$.  Whenever we perform a linear operation on all of the pixels at once, we can use the SVD $X = USV^T$ to reduce memory and computational costs by working with $US$ in place of $X$, which is in $N$ dimensions instead of $3WH \gg N$ (similar processing was done in \cite{turk1991eigenfaces} and \cite{tralie2017quasi}).  The only nonlinear operation is the median operation (Section~\ref{sec:cyclereordering}), and we project back into pixel space using $V^T$ before taking the median.

@@

where $\tau$ is a step size and $d$ is the dimension of the embedding.  In this work, we fix $\tau = 1$ for simplicity and drop $\tau$ in the subscript, and we also sample times $t$ only at integer frame indices.  As shown by Takens \cite{takens1981detecting}, a sliding window of dimension $d = 2m+1$ of even a single generic observation function (time series) of a dynamical system of intrinsic dimension $m$ is sufficient to reconstruct a topological embedding of the underlying trajectory in the original state space.  For periodic signals, this state space is a torus, and $d$ should be twice the number of harmonics present to injectively reconstruct loops on that torus \cite{perea2015sliding}.  The same is true of videos in general \cite{tralie2017quasi}.  Furthermore, the sliding window length $(d-1) \tau$, or simply $d-1$ in our case, maximizes the roundness of the embeddings when $d-1 = k T$ for some integer $k$, where $T$ is the fundamental period.  Since we will use topological data analysis (TDA) to find the scale of the loop, and since TDA measures roundness, it is important that we tune our window size to be an integer multiple of the period.  To do this, we perform 1D ISOMAP \cite{tenenbaum2000global} on the original frames of the video to generate a 1D ``surrogate signal,'' and we perform an autocorrelation-based fundamental frequency estimation \cite{Mcleod05asmarter} on this signal to estimate the period and, hence, the appropriate $d$.  This is quite similar to the diffusion maps + autocorrelation approach in \cite{tralie2017quasi} for analyzing periodic videos.  This estimation often returns integer multiples of the fundamental period, which is fine for our purposes.

@@

TDA addresses problems of scale

@@

Talk about how it can pick up on harmonics (e.g. jumping jacks), which is why this is important

@@

@@

Interpolated windows can be projected back to pixel space one by one to save memory

@@

\subsection{Robustness Tests}
Robustness experiments

\subsection{Qualitative Results}
\begin{figure}[h!]
\centering
\includegraphics[width=\columnwidth]{XTSliceJumpingJacks.pdf}
\caption{An XT slice of a line of pixels (magenta line, upper left) over time for an input video of two men doing jumping jacks and for reordered videos with and without median consensus.}
\label{fig:XTSliceJumpingJacks}
\end{figure}

\begin{figure}[h!]
\centering
\includegraphics[width=\columnwidth]{XTSliceFan.pdf}
\caption{An XT slice of a line of pixels (magenta line, upper left) over time for an input video of a fan with only 6 frames per period and the corresponding median consensus reordered video.}
\label{fig:XTSliceFan}
\end{figure}

Figure~\ref{fig:XTSliceJumpingJacks} shows the difference between a simple reordering and a median consensus reordering.  Due to natural variation from cycle to cycle, the simple reordering has many temporal discontinuities when interleaving these cycles.  By contrast, the median voting is clean, and it has the added benefit of removing nonperiodic background components.

umping jacks 2 men, two exercise videos from \cite{levy2015live}, beating heart videos from \cite{traliehigh}, \cite{wu2012eulerian}, and \cite{wadhwa2013phase}.
